---
title: 'Online Video Challenge'
author: "Siddhartha Jetti"
date: "7/4/2019"
output: rmarkdown::github_document
---

# Goal

The company of this challenge allows users to upload videos online, just like YouTube.

This company is interested in knowing whether a video is “hot” (i.e. trending up in terms of popularity), stable or going down. Understanding this would allow to optimize the videos promoted on the home-page and, therefore, maximize ads revenue.

# Challenge Description

Company XYZ is an online video streaming company, just like YouTube or Dailymotion.

The Head of Product has identified as a major problem for the site a very high home page drop-off rate. That is, users come to the home-page and then leave the site without taking any action or watching any video. 
Since customer acquisition costs are very high, this is a huge problem: the company is spending a lot of money to acquire users who don’t generate any revenue by clicking on ads.

Currently, the videos shown on the home page to new users are manually chosen. The Head of Product had this idea of creating a new recommended video section on the home page.

He asked you the following:

1) Classify each video into one these 3 categories: 
   ”Hot” - means trending up. These videos are candidate to be shown.
   “Stable and Popular” - video view counts are flat, but very high. These videos are candidates to be shown too.
   “Everything else” - these videos won’t be shown.
What are the main characteristics of the “hot videos”?

2) After having identified the characteristics of the hot videos, how would you use this information from a product standpoint?

# Data

We have 2 tables downloadable by clicking here.

The 2 tables are:

video_count - provides information about how many times each video was seen each day.

## Columns:

* video_id : video id, unique by video and joinable to the video id in the other table
* count : total count of views for each video
* date : on which day that video was watched that many times

video_features - characteristics of the video.

## Columns:

* video_id : video id, unique by video and joinable to the video id in the other table
* video_length : length of the video in seconds
* video_language : language of the video, as selected by the user when she uploaded the video
* video_upload_date : when the video was uploaded
* video_quality : quality of the video. It can be [ 240p, 360p, 480p, 720p, 1080p]

# Problem Setup

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(lubridate)
library(randomForest)


# Read in the input data into a dataframe
views <- read.csv("video_count.csv", stringsAsFactors = F) 
features <- read.csv("video_features.csv", stringsAsFactors = F) 
```

# Question 1:

Explore the views dataset
```{r}
# Transform views dataset
views <- views %>%
  mutate(date = as.Date(date)) %>%
  arrange(video_id, date)

# Check data types of each of the columns
summary(views)

# Take a peek at data
head(views)
```

Explore the features dataset
```{r}

# Transform features dataset
features <- features %>%
  mutate(video_upload_date = as.Date(video_upload_date)) %>%
  arrange(video_id)

# Check data types of each of the columns
summary(features)

# Take a peek at data
head(features)
```

Check for missing values in the views and features data.
```{r}
# count of missing values by column in views dataset
colSums(is.na(views))
# count of missing values by column in features dataset
colSums(is.na(features))
```
No missing values exist anywhere in the data.

Check if count data exists for all the videos for all the dates.
```{r}
table(views$date)
```

Clearly, The video counts are provided for 15 days starting from January 1st 2015.

```{r}
length(unique(views$video_id))
```

There are 2785 videos and 15 days of view counts data exist for every video.

Here the goal is to classify videos into Hot, stable and others based on if videos are trending up or down. Extracting trend from the time series of views count is essential to accomplish this task.

There are several ways to extract the trend. The simplest method is to fit a straight line that is closest to the views count data points for each video, In other words finding the line that minimizes the total sum of squared distances between the points and line. 

Let's pick few arbitrary videos and visualize the time series of views.

```{r}
views %>%
  filter(video_id == 176) %>%

ggplot(aes(date, count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
views %>%
  filter(video_id == 499) %>%

ggplot(aes(date, count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

For the selected video "176", the least sum of squares line appears to be a good fit to the data. But for video "499", it looks like a poor fit. Since the goal here is not to predict future counts but only for a descriptive purpose, the linear trend line approach can still be used. The slope of trendline is the coefficient of date variable in regression line equation. The positive slope indicates that views are trending up and a negative slope indicates the video is trending down in terms of views over time,

Before finding the least sum of squares fit to all the videos, let us normalize the variables so that visualizing the trendlines of various videos becomes easy.

```{r}
# Function to normalize the variables
normalize <- function(x) {
    return((x - min(x)) / (max(x) - min(x)))
}

# Function to extract slope and trend in the data
slope_trendline <- function(y, x) {
  trendline <- lm(formula = y ~ x)
  return(trendline$coefficients[2])
}

# Now group by video id, normalize variables and extract slope
views_summary <- views %>%
  mutate(date = as.numeric(date),
         count_norm = normalize(count),
         date_norm = normalize(date)) %>%
  group_by(video_id) %>%
  summarise(avg_views_per_day = round(mean(count), digits = 0),
            slope = slope_trendline(count_norm, date_norm))

# Take a look at the data
head(views_summary)
```

Here is the distribution of average counts per day 
```{r}
quantile(views_summary$avg_views_per_day, probs = seq(0, 1, by = 0.05)) 
views_summary %>%
  ggplot() +
  geom_histogram(bins = 30, aes(x = avg_views_per_day, y = ..density..)) +
  geom_density(aes(x = avg_views_per_day, y = ..density..))
```

The distribution of slopes of trendlines
```{r}
quantile(views_summary$slope, probs = seq(0, 1, by = 0.1)) 
views_summary %>%
  ggplot() +
  geom_histogram(bins = 30, aes(x = slope)) 
```

Based on the above distributions, Here are the rules that can be thought of to classify videos into Hot, stable and others.

* If slope >= 0.02 then video is "Hot"
* If -0.02 < slope < 0.02 and average views per day more than 1,000,000  then  video is "Stable and Popular".
* If none of the above then video is "Everything else""

```{r}
# Classifying videos based on the stated rules
views_summary <- views_summary %>%
  mutate(category = case_when(
                      slope >= 0.02 ~ "Hot",
                      slope > -0.02 & slope < 0.02 & avg_views_per_day > 1000000 ~ "Stable and Popular",
                      TRUE ~ "Everything else")
  )

# Frequency of different video categories
table(views_summary$category)
```

# Question 2:

Now, The goal is to build a model to predict if a video is "Hot" and understand the factors that effect it.

Run descriptive stats on the video features dataset
```{r}
# Merging the two datasets
data <- features %>%
  left_join(views_summary, by = "video_id") %>%
  mutate(is_hot = as.factor(ifelse(category == "Hot", 1, 0)),
         days_after_upload = as.Date("2015-01-15") - video_upload_date,
         video_language = as.factor(video_language),
         video_quality = as.factor(video_quality),
         upload_weekday = as.factor(weekdays(video_upload_date)),
         upload_week = week(video_upload_date)) %>%
  select(-video_upload_date, -category, -slope)

# Variable data types
summary(data)
```

```{r}
data %>%
  group_by(is_hot) %>%
  summarise(avg_length = mean(video_length)) %>%
ggplot() +
  geom_col(aes(x = is_hot, y = avg_length))

ggplot(data, aes(x = video_length, group = is_hot)) +
  geom_histogram(binwidth = 120, aes(fill = is_hot))
```

The average length of Hot videos is much lower than other videos. The histogram of video length reveals that proportion of Hot videos is much lower in the long videos than shorter ones.

Average views per day by language
```{r}
data %>%
  group_by(video_language) %>%
  summarise(avg_views_per_day_per_video = mean(avg_views_per_day)) %>%
ggplot() +
  geom_col(aes(x = video_language, y = avg_views_per_day_per_video))
```


Number of videos uploaded and proportion of hot videos by language
```{r}
data %>%
  group_by(video_language, is_hot) %>%
  summarise(counts = n()) %>%
  mutate(freq = round(counts / sum(counts), digits = 2)) %>%
ggplot(aes(x = video_language, y = counts, group = is_hot)) +
       geom_col(aes(fill = is_hot)) +
       geom_text(aes(label = paste0(freq*100, "%")))
```

The bar chart reveals that German and French videos are much less in number than other languages. The highest number of videos exist in English and Chinese than other languages.

Number of videos uploaded and proportion of hot videos by quality
```{r}
data %>%
  group_by(video_quality, is_hot) %>%
  summarise(counts = n()) %>%
  mutate(freq = round(counts / sum(counts), digits = 2)) %>%
ggplot(aes(x = video_quality, y = counts, group = is_hot)) +
       geom_col(aes(fill = is_hot)) +
       geom_text(aes(label = paste0(freq*100, "%")))
```

Number of videos uploaded and proportion of hot videos by upload week day
```{r}
data %>%
  group_by(upload_weekday, is_hot) %>%
  summarise(counts = n()) %>%
  mutate(freq = round(counts / sum(counts), digits = 2)) %>%
ggplot(aes(x = upload_weekday, y = counts, group = is_hot)) +
       geom_col(aes(fill = is_hot)) +
       geom_text(aes(label = paste0(freq*100, "%")))
```

## Model Building:

Random forest would be a good choice for predicting "Hot" videos as it is strong with outliers and works well with correlated, continuous and discrete variables. Also, random forest is easier to optimize parameters. Partial dependence plots can be used to capture insights from the model.

A standard 66-30% split can be used to generate training and test datasets. If training dataset is not large enough then cross validation can be used to arrive at the optimum parameters

```{r}

# Split the available data into training and test data sets 
set.seed(2019)
train_sample = sample(nrow(data), size = round(nrow(data)*0.66))
train_data = data[train_sample,]
test_data = data[-train_sample,]

# Build Random forest model on the data with mostly default settings except for class weight and #trees
rf.fit <- randomForest(y = train_data$is_hot, x = train_data[, -c(5,6)], ytest = test_data$is_hot, 
                       xtest = test_data[, -c(5,6)],  ntree = 250, mtry = 3, keep.forest = TRUE)

rf.fit

# Visualize Important variables
varImpPlot(rf.fit)
```

```{r}
# Order variables by importance
imp_vars <- importance(rf.fit)
pdp_vars <- names(sort(imp_vars[,1], decreasing = T))

# Loop through variables and build PDP in the decreasing order of variable importance
for (i in seq_along(pdp_vars)) {
  partialPlot(rf.fit, data[,-c(5,6)], pdp_vars[i], xlab = "", 
              main = paste("Partial Dependence on", pdp_vars[i]), which.class = 1)
}

```

The variable importance plot reveals that video length is a major factor in predicting if the video is trending or not followed by days after upload and video quality variables. The variable upload week is just proxy for days after upload.

The following are the characteristics of videos that are trending up:

* Short video lengths.

* Newer videos are more likely to be trending than older ones. This is expected as videos age, they lose the novelty affect and become less effective at generating more views.

* Low resolution videos (240p) have a slight higher chance of trending on the platform than higher resolution. This could be because lower resolution videos are easier to stream in low band width areas like streaming videos with mobiles on buses and subways.

* German and French videos are more likely to be trending than other languages. This is not suprising as there are so few German and French videos uploaded on the site and the low supply in those languages could be resulting in higher proportion of Hot videos.

# Question 3:

Let us dig deep and explore relationship between video length and views.
```{r}
# Bin vodeos based on the video length
data <- data %>%
  mutate(video_length_type = case_when(video_length > 0 & video_length <= 300 ~ "0-5 mins",
                                  video_length > 300 & video_length <= 600 ~ "05-10 mins",
                                  video_length > 600 & video_length <= 900 ~ "10-15 mins",
                                  TRUE  ~ "15+ mins")
         )

data %>%
  group_by(video_length_type) %>%
  summarise(avg_views_per_day_per_video = mean(avg_views_per_day)) %>%
ggplot() +
  geom_col(aes(x = video_length_type, y = avg_views_per_day_per_video))
```
 
The above plot shows that long videos (15+ mins duration) are generating far lower number of views than short ones (less than 10 min duration) and average views per day is lower by more than 35%. Also,The partial dependence plots reveal that short videos are more likely to be trending than longer ones. 

From the product stand point, Here are some recommendations:

* The company should investigate why there are so few German and French videos uploaded. This could be due to some translation problems in the upload process or is it something else.

* Clearly, it appears that lengthy videos are generating far fewer views and are less likely to be trending than short ones. So, The Company should encourage video creators to make short ones with engaging content. They could consider creating a seperate "short video" section with a limit on duration of video being uploaded somewhere between 5-10 mins. The company should commission studies on what the optimum allowable video length should be for the short video section.

* The company could also consider an approach to prompt users with a pop up telling the user that videos shorter than 10 mins are more likely to generate more views. If user is uploading really long videos then forbidding from uploading video to platform.
