---
title: "Diversity in the Workplace"
author: "Siddhartha Jetti"
date: "May 30, 2019"
output: rmarkdown::github_document
---

# Goal

Diversity, unconscious bias in the workplace and, in general, the way companies treat their employees are a very important topic. 
Data science can help discover potential discriminations by looking at the data and see if there are segments of employees that are treated worse.

# Challenge Description

There has been lots of talking about diversity in the workplace, especially in technology. The Head of HR at your company is very concerned about that and has asked you to analyze internal data about employees and see whether results suggest that the company is treating all its employees fairly or not.

Specifically, she gave you the following tasks:

1) In the company there are 6 levels (described below). Identify, for each employee, her corresponding level. 
* Individual Contributors(IC) - they don’t manage anyone
* Middle Managers(MM) - they are the direct bosses of IC
* Directors(D) - they are the direct bosses of MM
* VP - D direct bosses
* Executives (E) - VP direct bosses
* CEO - The direct boss of E.

2) How many people each employee manages? Consider that if John directly manages 2 people and these two people manage 5 people each, then we conclude that John manages 12 people.


3) Build a model to predict the salary of each employee.
Describe the main factors impacting employee salaries. Do you think the company has been treating all its employees fairly? What are the next steps you would suggest to the Head of HR?

# Data

We have 2 tables downloadable by clicking here.

The 2 tables are:

company_hierarchy - info about each employee direct boss and her dept

## Columns:

* employee_id : the Id of the employee. It is unique by employee and can be joined to employee id in the other table
* boss id : the id of the boss of employee id. It is unique by employee and can be joined to employee id in the other table.
* dept : employee id dept. There are the following departments: 
  + Engineering (data science is under engineering)
  + Marketing
  + Sales
  + HR
  + the “CEO” dept just for the CEO, since belongs to all the above dept.

employee - info about each employee

## Columns:

* employee_id : the Id of the employee. It is unique by employee and can be joined to employee id and/or boss id in the other table
* signing_bonus : whether the employee got a signing bonus when she joined the company (1 -> yes, 0 -> no)
* salary : the current salary of that employee in USD
* degree level: the highest degree received by the employee.
* sex: Male/Female 
* yrs_experience: employee years of work experience


# Problem Setup

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(randomForest)

# Read in the input data into a dataframe
data_hierarchy <- read.csv("company_hierarchy.csv")
data_employee <- read.csv("employee.csv")
```


# Question 1:

Get the count of reports and level from hierarchy table.  
```{r}
hierarchy <- data_hierarchy %>%
  filter(!is.na(boss_id))

# Function to count direct/in-direct reports and get the hierarchy level for every employee
countReports <- function(df,id) {
  reports <- df[df$boss_id == id,]
  reports1 <- reports
  len <- nrow(reports)
  level <- ifelse(len == 0, 1, 2)
  while(1){
    reports <- df[df$boss_id %in% reports$employee_id,]
    reports1 <- rbind(reports1, reports)
    if(nrow(reports) == len){
     return(data.frame(employee_id = id, level = level, report_count = nrow(reports1)))
     break
    } 
    len <- nrow(reports)
    level <- level + 1
  }
}

data_reports <- data.frame(employee_id = NULL, level = NULL, report_count = NULL )
emps <- nrow(data_hierarchy)

# Loop through all the employess to get level and reports count.
for(i in 1:emps){ 
  data_reports <- rbind(data_reports, countReports(hierarchy, data_hierarchy$employee_id[i]))
}

# Join hierarchy related information with employee information to form one table
data <- data_reports %>%
  arrange(desc(report_count), employee_id) %>%
  left_join(data_hierarchy, by = "employee_id") %>%
  select(employee_id, dept, level, report_count) %>%
  right_join(data_employee, by = "employee_id") %>%
  mutate(signing_bonus = as.factor(signing_bonus),
         level = as.factor(case_when( level == 3 ~ "MM",
                            level == 4 ~ "D",
                            level == 5 ~ "VP",
                            level == 6 ~ "E",
                            level == 7 ~ "CEO",
                            TRUE ~ "IC" ))
  )
```

# Question 2:

Run descriptive statistics 
```{r}
# Check datatypes of all the variables in the dataframe
str(data)
```

```{r}
# Take a peek at the data
head(data)
summary(data)

# check for duplicate user ids
length(data$user_id) == length(unique(data$user_id))
```

Here are some observations about data:

* There are total of 10000 employees in dataset and each employee id is unique
* The mean and median salary of employess across the company are USD 182k and 189k respectively.

```{r}
# Median is more robust to outliers than mean
salary_dept = data %>%
                filter(!dept %in% c("CEO")) %>%
                group_by(dept) %>%
                summarise(med_salary = median(salary))
ggplot(data = salary_dept, aes(x = dept, y = med_salary))+
       geom_bar(stat = "identity")

```

It looks like folks in engineering are earning way more than HR and other depts. This could mean that there is a bias by dept or it could just be a spurious correlation, which would be known only after building a model.

```{r}
salary_level = data %>%
                group_by(level) %>%
                summarise(med_salary = median(salary)) %>%
                arrange(desc(med_salary))

ggplot(data = salary_level, aes(x = level, y = med_salary))+
       geom_bar(stat = "identity")
```
The above plot indicates that median salaries of CEO > Executive > VP > Director > MM > IC which makes sense given the increasing scope and responsibility.

```{r}
salary_degree = data %>%
                group_by(degree_level) %>%
                summarise(med_salary = median(salary)) %>%
                arrange(desc(med_salary))

ggplot(data = salary_degree, aes(x = degree_level, y = med_salary))+
       geom_bar(stat = "identity")
```

Overall, median salary for bachelor degree graduates is slightly lower than highschool and increases with advanced degrees. PHD graduates have the highest median salary. One possible explanation for lower bachelor graduate salary than high school graduates could be that gains in salary because of higher degree are nullified by loss of several yrs of experience (4 or more years).

```{r}
salary_gender = data %>%
                group_by(sex) %>%
                summarise(med_salary = median(salary)) %>%
                arrange(desc(med_salary))

ggplot(data = salary_gender, aes(x = sex, y = med_salary))+
       geom_bar(stat = "identity")
```

The plot shows that median salary of Males is significantly higher than females. However, This plot alone doesn't confirm the bias as gender could be a proxy for other variables like educational degree, year of experience etc. that are known to determine the salary.

```{r}
# Plot salary vs yrs of experience
ggplot(data = data, aes(x = yrs_experience, y = salary)) +
  geom_point() + 
  geom_smooth(method = "loess", se = FALSE)
  
```

The plot indicates that salaries are expected to increase with yrs of experience and are expected to grow faster after 10-15 yrs of experience.

```{r}
# Plot salary vs number of reports
data %>%
  filter(report_count > 0) %>%
ggplot(aes(x = report_count, y = salary)) +
  geom_point() + 
  geom_smooth(method = 'loess', se = FALSE)
```

# Model Building 

Random forest would be a good choice for predicting salary as it is strong with outliers and works well with correlated, continuous and discrete variables. Also, random forest is easier to optimize parameters.Partial dependence plots can be used to capture insights from the model. 

A standard 66-30% split can be used to generate training and test datasets. If training dataset is not large enough then cross validation can be used to arrive at the optimum parameters

```{r}

# Training and test set split
train_sample = sample(nrow(data), size = round(nrow(data)*0.66))
train_data = data[train_sample,]
test_data = data[-train_sample,]

# Build Random forest model on the data with mostly default settings except for class weight and #trees
set.seed(2019)
rf.fit <- randomForest(y=train_data$salary, x = train_data[,-c(6)], ytest = test_data$salary, 
                       xtest = test_data[,-c(6)],  ntree = 250, mtry = 3, keep.forest = TRUE)

rf.fit
# Visualize Important variables
varImpPlot(rf.fit)
```
The training and test MSE are close and looks like model is not overfitting to data.
Now, Building the model without the dominating variable to see if it changes importance of other variables.

```{r}
rf.fit2 <- randomForest(y=train_data$salary, x = train_data[,-c(2,6)], ytest = test_data$salary, 
                       xtest = test_data[,-c(2,6)],  ntree = 250, mtry = 3, keep.forest = TRUE)

rf.fit2
# Visualize Important variables
varImpPlot(rf.fit2)
```
After dropping the dept variable, the variance explained drops dramatically. So rf.fit is indeed optimal model.

# Question 3:

```{r}
# Order variables by importance
imp_vars <- importance(rf.fit)
pdp_vars <- names(sort(imp_vars[,1], decreasing = T))

# Loop through variables and build PDP in the decreasing order of variable importance
for (i in seq_along(pdp_vars)) {
  partialPlot(rf.fit, data[,-6], pdp_vars[i], xlab="",
              main=paste("Partial Dependence on", pdp_vars[i]))
}

```

# Understanding Partial Dependence Plots

* Looking at the variable importance and PD plots, the main factors that impact the salaries are dept, yrs of experience and level or number of direct/indirect reports.

* Although the descriptive bar chart of Salary Vs Gender show that Females are underpaid compared to Males, after controlling for variables like department, yrs of experience and number of reports no such bias in salary based on gender exists.

* The PDP on dept reveals that salaries in engineering are higher than any other department and salaries in HR are the lowest.

* Interestingly after controlling for dept, yrs of experience and level, educational degree does not really affect the salary.

* Signing bonus and gender does not seem to affect the salary. 

# Conclusions & next steps

* Based on the data, the people working in engineering dept are likely to get higher salaries than other depts. However, the reason for higher salaries could be long work hours, high stress levels, scarcity of engineering talent or difficulty in retaining talent etc.

* The next steps for the Head of HR would be to identify the most likely reason and try to reduce the salary gap across the depts.

